Submodular Optimization for Control of Prosumer Networks

### 2016-04-07 Article.tex
corrections Monique and Michel

\begin{document}
\pagestyle{plain}
\title{Control of Prosumer Networks}

\author{Nicolas Gensollen, Monique Becker, Vincent Gauthier and Michel Marot  \\
\IEEEauthorblockA{CNRS UMR 5157 SAMOVAR, \\
Telecom SudParis/Institut Mines Telecom\\
Email: \{nicolas.gensollen, vincent.gauthier, michel.marot, monique.becker\}@telecom-sudparis.eu}}

@@

In this paper, we study the case of a network composed of entities called prosumers. These agents have the ability to both consume and produce according to external conditions. We first model and simplify the underlying dynamic of such network using a second order coupled oscillators network model. Under some conditions, the system synchronize to a common frequency. However, in case of perturbations in the power distribution, the system might lose synchrony and require control to bring it back to the stable state. In this situation, control can be seen as energy  absorbed or injected in the system at specific locations. Moreover, the power outputs of the prosumers are susceptible to change such that loads and generators are not fixed. In this context, it is important to select the right subset of nodes that yields, on average, the cheapest control in terms of energy. 
\end{abstract}

@@

This paper is organized as follows, section \ref{sec:control} presents some basic notions about control in networks, section \ref{sec:The_Model} introduces the oscillator model used for simplifying the grid's dynamics. In section \ref{sec:control_and_submodularity} we explain how we select the subset of controlled nodes such that the average control energy is optimized. In section \ref{sec:constraints} we develop the constraints relative to the system and the batteries. Finally, in section \ref{sec:Results}, we show some results.

@@

If there is no action from the outside, the system state will evolve according to $ X(t) = X_0 e^{At} $ where $X_0$ is the initial state of the system. Imagine now, that this dynamic could be somehow influenced by injecting some signals. More precisely, lets assume that we can inject these signals in a subset of the nodes such that the dynamics becomes $ \dot{ X }(t) = A X(t) + B u(t) $. Where the matrix B indicates which nodes receive the signals and $ u(t) $ is the vector of inputs at time t. Assume that the system is in some initial state $X_0$ and we aim at bringing it to some final state $X_T$ in some amount of time T. Control can be seen as finding the sequence of inputs $\{u(t_0),...u(t_T)\} $ that would realize such a thing given the dynamics A. Furthermore, we say that the system is controllable in T steps if it can be steered from any initial state $ X_0 $ to any final state $ X_T $ through a sequence of control inputs.

It was shown by Kalman in 1963 that a linear system $(A,B)$ is controllable if the $N \times NM $ controllability matrix $ C = [B,AB,A^2B,...,A^{N-1}B] $ has full rank, that is, $rank(C) = N$. Nevertheless, when studying the controllability of non trivial networks, using the Kalman's criterion poses some difficulties since matrix C becomes huge. Furthermore, because $A^k\longrightarrow 0$ when $k\longrightarrow \infty $, numeric precision becomes quickly an issue when computing the rank of C. A different approach to study controllability in complex network was introduced later on by Lin \cite{1100557}. This relates the search of the minimum driver set to a maximum matching in a bipartite graph, which can be done with the Hopcroft-Karp algorithm in $ O(|E|\sqrt{|V|}) $ in the worst case. This requires that the system $(A,B)$ is a structured system, meaning that all elements in A and B are either fixed zeros or independant free parameters. In such a situation, structural control is extremely powerfull since it enables us to discuss a network's controllability relatively quickly even if we do not know the exact weights of the edges. Note however that the independance of the parameters is a key assumption. If A is the adjacency matrix of an undirected network for instance, independance is not verified (since A is symetric). In this case, one should use different tools to study controllability \cite{Yuan2014}.

On the other hand, if the system is controllable, there might be more than one sequence of control inputs $u(t)$ that could drive it from $X_0$ to $X_T$. Among all these possibilities, optimal control is devoted to find the sequence that minimizes some cost function that can depend on the sates of the system $X(t)$, the control inputs $u(t)$, and the final state $X_T$. In the following, these control inputs will be used to model the actions of storage devices in some power grid. We are therefore particularly interested in the sequence that requires the minimum amount of control energy \cite{Yan2012}. Let $ \mathcal{E} = \int_{0}^{T} \Vert u(t) \Vert^2 dt $ be the energy used for the control of the system. It can be shown \cite{Liu2015} that the control input sequence that minimizes $ \mathcal{E} $ can be written as :

@@

\section{Power Grid Dynamics}
\label{sec:The_Model}

@@

Since the dynamics is expressed in term of deviations from $ \Omega$, synchronization at $ \Omega $ is achieved if $\omega_{SYNC} = 0 $. Wich gives the production consumption balance constraint :

@@

As for power lines, we need to take into account the imperfections of the storage that we consider. In practice, we could use some detailed and complex models, but as to keep things simple, we adopt here a straightforward approach. A storage i has a maximum charge/discharge rate $r_i$, some amount of energy stored at time t $ \Lambda_i(t) $, and a maximum capacity $ \Lambda_{i, MAX} $. We assume that all maximum rates ($r$) and all maximum capacities $\Lambda_{Max}$ are the same accross the storage equipments. We also denote by $ \Lambda(t) $ the vector of energy level at time t. Since $ u^{\star}(t) $ specifies the control inputs, the energy level dynamic of the batteries is $ \Lambda(t+\Delta t) = \Lambda(t) - u^{\star}(t) I \Omega $. Which can also be written as :

@@

Finally, during each time slot, the battery cannot charge or discharge at a rate higher than $ r $ :

@@

\begin{figure}
\includegraphics[scale=.55]{plot5.pdf}%./figure_2/figure_2.pdf}
\caption{At time $ t_k=4500 $, $P_i(t_k) = P_i(t_{k-1}) + \delta P$ for some random node i (perturbation). Consequently, the frequencies $ \dot{ \theta }_i $ deviate from the synchronized state. At time $ t_k + d = 5000$, optimal control inputs (see equation \ref{eq:ustar}) are injected such that the system is brought to the synchronized state at time $ t_k + d + T = 5500 $ (T is the control time).}

@@

Finding the driver set of size k $S_k$ using the gramian $W_{S_k}$ of the system $(A,B_{S_k})$ does not require initial and final states. Besides, $W_{S_k}$ does not depend on the modified power distribution $ \Psi $ of the nodes. We are indeed looking for k controllers that perform well on average over all possible situations. On the contrary, the constraints derived above (eq. \ref{eq:flow_cons}, \ref{synchro_constraint}, \ref{level_constraints} and \ref{rate_constraints}) are bound to a particular trajectory in the state space. We indeed check wether the power grid can sustain the controlled dynamics when transitioning from a given initial state to a target final state. As we do not know what these states could be, we could use multiple scenarios and test wether $S_k$ can control the system without violation of constraints. In this paper, we use a slightly different approach. We consider that the system is initially at equilibrium (all elements are synchronized at $\Omega$) and that control will be necessary if a perturbation (power inbalance) occurs and takes the system out of equilibrium (see figure \ref{fig:frequecies}). Depending on how much time we need to start the control phase, the state of the system (the initial state for control) might be "somewhere around" the synchronized state. In order to test wether a set $S_k$ can control the system without violation of constraints, we sample initial states within some hypersphere centered on the synchronized state and check all the constraints. If for all trajectories, the constraints are respected, then we consider that $S_k$ enables the control of the system.

@@

The purpose of selecting the drivers according to a Gramian related metric is to minimize, on average, the amount of control energy needed. In other words, if we select the drivers randomly, we expect to need, on average, more energy to control the system. In figure \ref{fig:densities}, we draw $10^4$ example systems of 50 nodes with random scale-free topologies (Barabasi-Albert model), and random power distributions and line capacities.  For each system we select a random number of drivers $N_D \sim \mathcal{U}(1,N_{nodes}) $ and we find two driver sets of size $N_D $. One is chosen randomly and the other is found with algorithm \ref{algo_1}. For both sets, if the control is possible, we draw a random initial state $Y_i$ and a random final state $Y_f$, and we compute the control energy required for driving the system from $Y_i$ to $Y_f$. On figure \ref{fig:densities} we plot each system as a dot with coordinates $(n_D = N_D / N_{nodes}, log(\mathcal{E}))$. The color indicates what method has been used, and the lines show the average accross $n_D$. As visible, both curves decrease as the number of drivers increases, meaning that, as the number of drivers grows, the average control energy tends to decrease. Furthermore, we tend to use, as expected, less energy when the drivers are carefully selected. Note that this difference tends to zero when $n_D$ tends to one, because almost all nodes are then selected, yielding little flexibility for optimization.

@@

\begin{figure}
\includegraphics[scale=.55]{plot4.pdf}%./figure_3/figure_3}
\caption{Logarithm of average control energy $log(\mathcal{E})$ against size of control set $n_D$ for random and optimized driver sets. Topologies are drawn from a Barabasi-Albert model with $ N_{nodes} = 50 $, power and line capacity distributions are random. $10^4$ instances are considered and the results are displayed as dots. The lines shows the average for both case. }
\label{fig:densities}
\end{figure}

We investigate next how the topology of the grid and the physical constraints affect the size of the driver set. We consider the simple case of an erdos renyi topology with probability of connection p. We show on Figure \ref{fig:erdos_renyi} how the size of the driver set evolves with p for different Gramian based metrics and for two levels of constraints :

\begin{itemize}
\item Level 1 : full control only : the system can be moved from any point to any other point of the state space. 
\item Level 2 : full control and grid constraints (see section \ref{sec:control_and_submodularity}). The system can be moved from any point to any other point without overloading any line or breaking any battery limits in the process.
\end{itemize}

\begin{figure}
\includegraphics[scale=.55]{plot3.pdf}%./figure_4/figure_4}
\caption{ $ n_D $ against link probability p for erdos-renyi topologies ($ N=100 $). Curves are averaged over 100 realizations. The top three curves show the results for the three metrics taken into consideration when all constraints are considered (see Level 2 in the main text). The bottom three curves exhibit the results for the same metrics when only the full controllability constraint is considered (see Level 1).}
\label{fig:erdos_renyi}
\end{figure}

When $p \sim 0$, nodes tend to be very poorly connected such that we need to control almost all nodes in the grid. As p increases, the connectivity of the graph augments and the number of drivers decreases. At some point, the connectivity of the graph starts to harm its controllability, and more drivers are needed (this effect is in accordance with the litterature). As expected, the driver sets for the level 2 of constraints are larger than for level 1 (for all metrics) because we impose far more constraints on the control inputs.

>>>>

\begin{document}
\pagestyle{plain}
\title{Submodular Optimization for Control of Prosumer Networks}

\author{Nicolas Gensollen, Vincent Gauthier, Monique Becker and Michel Marot  \\
\IEEEauthorblockA{CNRS UMR 5157 SAMOVAR, \\
Telecom SudParis/Institut Mines Telecom\\
Email: \{nicolas.gensollen, vincent.gauthier, monique.becker, michel.marot\}@telecom-sudparis.eu}}

@@

In this paper, we study the case of a network composed of entities called prosumers. These agents have the ability to both consume and produce according to external conditions. We first design a model and simplify the underlying dynamic of such network using a second order coupled oscillators network model. Under some conditions, the system synchronizes to a common frequency. However, in case of perturbations in the power distribution, the system might lose synchrony and require control to bring it back to the stable state. In this situation, control can be seen as energy  absorbed or injected in the system at specific locations. Moreover, the power outputs of the prosumers are susceptible to change such that loads and generators are not fixed. In this context, it is important to select the right subset of nodes that yields, on average, the cheapest control in terms of energy. We propose and validate an algorithm based on a submodular optimization for chosing the driver nodes.
\end{abstract}

@@

This paper is organized as follows, section \ref{sec:control} presents some basic notions about control in networks, section \ref{sec:The_Model} introduces the oscillator model designed for simplifying the grid's dynamics. In section \ref{sec:control_and_submodularity} we propose how we select the subset of controlled nodes such that the average control energy is optimized. In section \ref{sec:constraints} we develop the constraints relative to the system and the batteries. Finally, in section \ref{sec:Results}, we show some results and validation.

@@

If there is no action from the outside, the system state will evolve according to $ X(t) = X_0 e^{At} $ where $X_0$ is the initial state of the system. Imagine now, that this dynamic could be somehow influenced by injecting some signals. More precisely, lets assume that we can inject these signals in a subset of the nodes such that the dynamics becomes $ \dot{ X }(t) = A X(t) + B u(t) $, where the matrix B indicates which nodes receive the signals and $ u(t) $ is the vector of inputs at time t. Assume that the system is in some initial state $X_0$ and we aim at bringing it to some final state $X_T$ in some amount of time T. Control can be seen as finding the sequence of inputs $\{u(t_0),...u(t_T)\} $ that would do it given the dynamics A. Furthermore, we say that the system is controllable in T steps if it can be steered from any initial state $ X_0 $ to any final state $ X_T $ through a sequence of control inputs.

It was shown by Kalman in 1963 that a linear system $(A,B)$ is controllable if the $N \times NM $ controllability matrix $ C = [B,AB,A^2B,...,A^{N-1}B] $ has full rank, that is, $rank(C) = N$. Nevertheless, when studying the controllability of non trivial networks, using the Kalman's criterion is not a simple task since matrix C becomes huge. Furthermore, because $A^k\longrightarrow 0$ when $k\longrightarrow \infty $, numeric precision becomes quickly an issue when computing the rank of C. A different approach to study controllability in complex network was introduced later on by Lin \cite{1100557}. This relates the search of the minimum driver set to a maximum matching in a bipartite graph, which can be done with the Hopcroft-Karp algorithm in $ O(|E|\sqrt{|V|}) $ in the worst case. This requires that the system $(A,B)$ is a structured system, meaning that all elements in A and B are either fixed zeros or independant free parameters. In such a situation, structural control is extremely powerfull since it enables us to discuss a network's controllability relatively quickly even if we do not know the exact weights of the edges. Note however that the independance of the parameters is a key assumption. If A is the adjacency matrix of an undirected network for instance, independance is not verified (since A is symetric). In this case, one should use different tools to study controllability \cite{Yuan2014}.

On the other hand, if the system is controllable, there might be more than one sequence of control inputs $u(t)$ that could drive it from $X_0$ to $X_T$. Among all these possibilities, optimal control is devoted to find the sequence that minimizes some cost function that can depend on the sates of the system $X(t)$, the control inputs $u(t)$, and the final state $X_T$. In the following, these control inputs will be used to model the actions of storage devices in a power grid. We are therefore particularly interested in the sequence that requires the minimum amount of control energy \cite{Yan2012}. Let $ \mathcal{E} = \int_{0}^{T} \Vert u(t) \Vert^2 dt $ be the energy used for the control of the system. It can be shown \cite{Liu2015} that the control inputs that minimize $ \mathcal{E} $ can be written as :

@@

\section{Model design for power grid dynamics}
\label{sec:The_Model}

@@

Since the dynamics is expressed in term of deviations from $ \Omega$, synchronization at $ \Omega $ is achieved if $\omega_{SYNC} = 0 $. Which gives the production consumption balance constraint :

@@

As for power lines, we need to take into account the imperfections of the storage that we consider. In practice, we could use some detailed and complex models, but as to keep things simple, we adopt here a straightforward approach. A storage i has a maximum charge/discharge rate $r_i$, some amount of energy stored $ \Lambda_i(t) $ at time t, and a maximum capacity $ \Lambda_{i, MAX} $. We assume that all maximum rates ($r$) and all maximum capacities $\Lambda_{Max}$ are the same accross the storage equipments. We also denote by $ \Lambda(t) $ the vector of energy level at time t. Since $ u^{\star}(t) $ specifies the control inputs, the energy level dynamic of the batteries is $ \Lambda(t+\Delta t) = \Lambda(t) - u^{\star}(t) I \Omega $. Which can also be written as :

@@

During each time slot, the battery cannot charge or discharge at a rate higher than $ r $ :

@@

\begin{figure}
\includegraphics[scale=.55]{plot5.pdf}%./figure_2/figure_2.pdf}
\caption{At $ t_k=800 $, $P_5 $ goes from $-1$ to $-2$ (power inbalance). Consequently, the frequencies $ \dot{ \theta }_i $ deviate from the synchronized state. At $ t_k + d = 1000$, optimal control inputs (see equation \ref{eq:ustar}) are injected at nodes 0, 2, and 4 (nodes with ellipses) such that the system is brought to the synchronized state at time $ t_k + d + T = 1500 $ (T is the control time).}

@@

Finding the driver set $S_k$ of size k using the gramian $W_{S_k}$ of the system $(A,B_{S_k})$ does not require initial and final states. Besides, $W_{S_k}$ does not depend on the modified power distribution $ \Psi $ of the nodes. We are indeed looking for k controllers that perform well on average over all possible situations. On the contrary, the constraints derived above (eq. \ref{eq:flow_cons}, \ref{synchro_constraint}, \ref{level_constraints} and \ref{rate_constraints}) are bound to a particular trajectory in the state space. We indeed check wether the power grid can sustain the controlled dynamics when transitioning from a given initial state to a target final state. As we do not know what these states could be, we could use multiple scenarios and test wether $S_k$ can control the system without violation of constraints. In this paper, we use a slightly different approach. We consider that the system is initially at equilibrium (all elements are synchronized at $\Omega$) and that control will be necessary if a perturbation (power inbalance) occurs and takes the system out of equilibrium (see figure \ref{fig:frequecies}). Depending on how much time we need to start the control phase, the state of the system (the initial state for control) might be "somewhere around" the synchronized state. In order to test wether a set $S_k$ can control the system without violation of constraints, we sample initial states within some hypersphere centered on the synchronized state and check all the constraints. If for all trajectories, the constraints are respected, then we consider that $S_k$ enables the control of the system.

@@

The purpose of selecting the drivers according to a Gramian related metric is to minimize, on average, the amount of control energy needed. Conversely, if we select the drivers randomly, we expect to need, on average, more energy to control the system. In figure \ref{fig:densities}, we compare the average control energy  $log(\mathcal{E})$ in function of $n_D = N_D/N_{nodes}$ for a set of drivers selected thanks to our algorithm with randomly selected drivers (in both cases overloading and battery limits constraints are satisfied). We draw $10^4$ scale-free topologies with 50 nodes, and random power distributions and line capacities.  For each system we select a random number of drivers $N_D \sim \mathcal{U}(1,N_{nodes}) $ and we find two driver sets of size $N_D $. One is chosen randomly and the other is found with algorithm \ref{algo_1}. For both sets, if the control is possible, we draw a random initial state $Y_i$ and a random final state $Y_f$, and we compute the control energy required for driving the system from $Y_i$ to $Y_f$. Since the algorithm is based on $ Tr[W^{-1}]$, we could plot $ Tr[W^{-1}]$ in function of $n_D$ but we prefer to validate our method by computing the actual average energy. As expected, both curves decrease as the number of drivers increases, meaning that, as the number of drivers grows, the average control energy tends to decrease, but we tend to use less energy when the drivers are selected with our algorithm. Note that this difference tends to zero when $n_D$ tends to one, because almost all nodes are then selected, yielding little flexibility for optimization.

@@

\begin{figure}
\includegraphics[scale=.55]{plot4.pdf}%./figure_3/figure_3}
\caption{$log(\mathcal{E})$ against $n_D$ for random and optimized driver sets ($ N_{nodes} = 50 $).}
\label{fig:densities}
\end{figure}

We investigate next how the topology of the grid and the physical constraints affect the minimum size of the driver set. The driver set size $n_D$ is no longer selected randomly, but minimized. We consider the simple case of an Erdös-Renyi topology with probability of connection p. We show on Figure \ref{fig:erdos_renyi} how the minimum size of the driver set evolves with p for different Gramian based metrics and for two levels of constraints :

\begin{itemize}
\item Level 1 : full control and grid constraints (see section \ref{sec:control_and_submodularity}). The system is controllable (i.e it can be moved from any point to any other point) and under the constraints (i.e it can be moved without overloading any line or breaking any battery limits).
\item Level 2 : full control only : the system can be moved from any point to any other point of the state space, without considering any overloading or battery limit constraints. 
\end{itemize}

\begin{figure}
\includegraphics[scale=.55]{plot3.pdf}%./figure_4/figure_4}
\caption{ $ n_D $ against link probability p for erdos-renyi topologies ($ N=100 $). Curves are averaged over 100 realizations. The top three curves show the results for the three metrics taken into consideration when all constraints are considered (see Level 1 in the main text). The bottom three curves exhibit the results for the same metrics when only the full controllability constraint is considered (see Level 2).}
\label{fig:erdos_renyi}
\end{figure}

When $p \sim 0$, nodes tend to be very poorly connected such that we need to control almost all nodes in the grid. As p increases, the connectivity of the graph rises and the number of drivers decreases. At some point, the connectivity of the graph starts to harm its controllability, and more drivers are needed (this effect is in accordance with the litterature). As expected, the driver sets for the level 1 of constraints are larger than for level 2 (for all metrics) because we impose far more constraints on the control inputs.
